Typing a linear \picalc

### 2020-04-07 paper/paper.tex
Go over the intro again. Still needs more work

We live in a concurrent world where any given state is often computed interactively by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and even construct such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
The \picalc{} is a formal language that abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data through communication channels.
Notably, it features \emph{channel mobility}: the channels themselves are first order constructs, and can therefore be transmitted too.
We define the syntax for the \picalc{} in \S \ref{syntax}.
We use dependent types to make this syntax well-scoped by construction: every unbound variable is accounted for in the type of the process using it.
\todo{de Bruijn}

In the \lambdacalc{} all computations are local and happen on syntactically contiguous terms; in the \picalc{} however computations need not to happen on syntactically contiguous terms.
One way to define a reduction relation modulo these syntactic minutiae is to define a structural congruence relation that acts as a quotient type.
A term $A$ can then be freely rewritten into a term $B$ as long as $A$ is structurally congruent to $B$.
We introduce this equivalence relation in \S \ref{structural-congruence}.

In the \lambdacalc{} function application is the main computational engine; in the \picalc{} it is inter-process communication what drives computation.
Two subprocesses can communicate anywhere in a process but under input and output, which must be respected to ensure the sequentiality in behavior.
We provide an operational semantics for the \picalc{} in \S \ref{operational-semantics}.
Every computational step is the result of communication; we take care to keep information on what channel the communication is happening on so that we can later use it to state the theorem for subject reduction (aka type preservation).

As with the \lambdacalc{}, the \picalc{} can be used to construct a wide variety of misbehaved processes.

The properties guaranteed by the linear picalc

why resource-aware typing

extensional typing rules for a given syntax and operational semantics

leftover typing

@@

Type constructors as the names for the typing rules.
Datatype constructor rules might have additional index-free synonyms for ease of reference.
Double rule for type-level definitions.
We omit universe levels for brevity.
$\type{TYPES}$ are in blue and uppercased (and their indices are subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red and variables and functions in black.
Some constructors are shared across types. They can however always be disambiguated through the type of the goal.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}

@@

  \inferrule
  {x : \Var_n}
  {\suc x : \Var_{\suc n}}
\end{mathpar}
\caption{Types of size $n$}
\label{var}
\end{figure}

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
As shown in Figure \ref{process}, we increase the \emph{depth} counter every time we create a new channel or receive some input.

@@

\begin{equation*} \label{pi-calc-syntax}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar using de Bruijn indices}
\label{process}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}

@@

\subsection{Structural Congruence}

Structural congruence is a congruent equivalence relation on processes.
Any two structurally congruent processes are strongly bisimilar: they are can follow each other's reduction steps \cite{}.
Figure \ref{struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_0 \; Q$ witnesses that index $0$ does not appear nor in the inputs nor in the outputs of process $Q$ --- it does so by traversing $Q$.
The function $lower \; 0 \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $0$.
Finally, $swap \; 0 \; P$ traverses $P$ (scoped under $\suc \suc n$) and swaps variable references $0$ and $\suc 0$.

@@

  \inferrule
  { }
  {\constr{scope-comm} : \new \new P \eqeq \new \new swap \; 0 \; P}
\end{mathpar}
\caption{Structural rewriting rules. Premises $P$, $Q$ and $R$ are of type $\Process_n$ where $n$ can be inferred.}
\label{struct-cong-base}
\end{figure}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen at any point in a process' recursive definition, and they are closed under reflexivity, symmetry and transitivity as shown in Figure \ref{struct-cong}.
In \S \ref{subject-congruence} we will prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then $Q$ is well-typed.
Specifically, in the case of transitivity we must prove that if $P$ is structurally congruent with $Q$ and $Q$ with $R$, and $P$ is well-typed, then so is $R$.
To do so, we will have to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show that the doubly recursive call terminates we index the equivalence relation $=$ by the type $\Rec$, which models the structure of the recursion.

@@

  \inferrule
  {eq_1 : P \eq{r} Q \\ \; eq_2 : Q \eq{s} R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq{\constr{two} \; r \; s} R}
\end{mathpar}
\caption{Structural rewriting rules lifted to a congruent equivalence relation indexed by a recursion tree.
  Premises $P$, $P'$, $Q$, and $R$ are of type $\Process_n$ where $n$ can be inferred.}
\label{struct-cong}
\end{figure}

\subsection{Operational Semantics}

Figure \ref{reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common variable.
The receiving process substitutes references to its most immediate variable with references to the variable sent by the process doing the output, and is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- but notably not under input nor output, otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the variable over which communication happens.
If this variable is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the variable index over which communication occurs to the type level.
Every time we come out of a binder we decrement this variable.
To do so, we make use of a $dec$ function that saturates at $\constr{internal}$ --- as opposed to an $inc$ function that would not.

@@

  \inferrule
  {eq : P = P' \\ red : P' \reduce{i} Q}
  {\constr{struct} \; eq \; red : P \reduce{i} Q}
\end{mathpar}
\caption{Operational semantics indexed by the channel over which reduction occurs.}
\label{reduction}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Resource-aware Typing System}

@@

\subsection{Multiplicities and Capabilities}\label{multiplicities}

@@

\subsection{Typing with Leftovers}\label{leftover typing}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subject Reduction}\label{subject-reduction}

@@

\subsection{Subject Congruence}\label{subject-congruence}

>>>>

We live in a concurrent world where any given state is often computed interactively by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and even construct such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
The \picalc{} \cite{} is a formal language that abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data through communication channels.
Notably, it features \emph{channel mobility}: the channels themselves are first order constructs, and can therefore be transmitted too.
We type the syntax for the \picalc{} in \S \ref{syntax}.
We use de Bruijn indices \cite{} as variable references, and dependent types to make the syntax well-scoped by construction: every unbound variable is accounted for in the type of the process that uses it.

In the \lambdacalc{} all computations are local and happen on syntactically contiguous terms; in the \picalc{} however communicating subprocesses can be non-contiguous and even have different parent processes.
One way to define a reduction relation modulo these syntactic minutiae is to define a structural congruence \cite{} relation that acts as a quotient type.
A term $A$ can then be freely rewritten into a term $B$ as long as $A$ is structurally congruent to $B$.
We type this equivalence relation in \S \ref{structural-congruence}.

In the \lambdacalc{} function application is the main computational engine; in the \picalc{} it is inter-process communication what drives computation.
Two subprocesses can communicate anywhere in a process but under input and output, which must be respected to ensure the sequential behavior.
We type a reduction relation that acts as an operational semantics modulo subject congruence in \S \ref{operational-semantics}.
The base case of every such reduction is communication over a channel; we take care to track this information at the type level so that we can then use it to state the theorem for subject reduction (aka type preservation).

As with the untyped \lambdacalc{}, the untyped \picalc{} can be used to construct a wide variety of ill-behaved terms --- for some characterization of ill-behaved.
One can use type systems to constructively obtain a subset of terms for which certain properties hold.
As an example, the simply typed \lambdacalc{} ensures that the argument supplied to a function is of the expected type.
Correspondingly, the simply typed \picalc{} \cite{} ensures that the data sent over a channel is of the expected type.
This property is often referred to as \emph{communication safety}.

Another property of particular interest to concurrent systems is the avoidance of race conditions.
A type system that excludes multiple usages of a resource excludes race conditions.
In the linear \picalc{} \cite{} channels must be used exactly once.
This not only avoids races, but ensures \emph{communication privacy}: if both endpoints of a channel are acquired no other process can eavesdrop.
It also ensures that every declared communication channel must be used.
The linear \picalc{} is also interesting because together with product types it is isomorphic to session types \cite{} -- which are used to encode communication protocols.
We provide a typing system for the linear \picalc{} in \S \ref{type-system} as extensional rules on sitting on top of the unruly well-scoped syntax.

Linear type systems need of context splits.
These are tricky to handle constructively without explicitly asking for them to the user.
Leftover typing \cite{} is a type theoretical technique that addresses this.
It does so by carrying both an input and an output context (the leftovers).
The output context can then be used as the input context of the continuations.
We apply this technique to the \picalc{} in \S \ref{leftover-typing}.
We generalize our algebra on usage multiplicities in \S \ref{multiplicities}.
As a result, we are able to model a mix of shared, linear, and gradual types.
\todo{reformulate paragraph}

\todo{subject reduction}

@@

Data type definitions have double lines and index-free synonyms as rule names, for ease of reference.
Otherwise, we use the constructor name as the name of a typing rule.
We omit universe levels for brevity.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructors can be shared across types and disambiguated by context.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}

@@

  \inferrule
  {x : \Var_n}
  {\suc x : \Var_{\suc n}}
\end{mathpar}
\caption{Types of size $n$}
\label{fig:var}
\end{figure}

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
As shown in Figure \ref{fig:process}, we increase the \emph{depth} counter every time we create a new channel or receive some input.

@@

\begin{equation*}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar using de Bruijn indices}
\label{fig:syntax}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}
\label{semantics}

@@

\subsection{Structural Congruence}
\label{structural-congruence}

Structural congruence is a congruent equivalence relation on processes.
Any two structurally congruent processes are strongly bisimilar: they are can follow each other's reduction steps \cite{}.
Figure \ref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_0 \; Q$ witnesses that index $0$ does not appear nor in the inputs nor in the outputs of process $Q$ --- it does so by traversing $Q$.
The function $lower \; 0 \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $0$.
Finally, $swap \; 0 \; P$ traverses $P$ (scoped under $\suc \suc n$) and swaps variable references $0$ and $\suc 0$.

@@

  \inferrule
  { }
  {\constr{scope-comm} : \new \new P \eqeq \new \new swap \; 0 \; P}
\end{mathpar}
\caption{Structural rewriting rules. Premises $P$, $Q$ and $R$ are of type $\Process_n$ where $n$ can be inferred.}
\label{fig:struct-cong-base}
\end{figure}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen at any point in a process' recursive definition, and they are closed under reflexivity, symmetry and transitivity as shown in Figure \ref{fig:struct-cong}.
In \S \ref{subject-congruence} we will prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then $Q$ is well-typed.
Specifically, in the case of transitivity we must prove that if $P$ is structurally congruent with $Q$ and $Q$ with $R$, and $P$ is well-typed, then so is $R$.
To do so, we will have to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show that the doubly recursive call terminates we index the equivalence relation $=$ by the type $\Rec$, which models the structure of the recursion.

@@

  \inferrule
  {eq_1 : P \eq{r} Q \\ \; eq_2 : Q \eq{s} R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq{\constr{two} \; r \; s} R}
\end{mathpar}
\caption{Structural rewriting rules lifted to a congruent equivalence relation indexed by a recursion tree.
  Premises $P$, $P'$, $Q$, and $R$ are of type $\Process_n$ where $n$ can be inferred.}
\label{fig:struct-cong}
\end{figure}

\subsection{Operational Semantics}
\label{operational-semantics}

Figure \ref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common variable.
The receiving process substitutes references to its most immediate variable with references to the variable sent by the process doing the output, and is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- but notably not under input nor output, otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the variable over which communication happens.
If this variable is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the variable index over which communication occurs to the type level.
Every time we come out of a binder we decrement this variable.
To do so, we make use of a $dec$ function that saturates at $\constr{internal}$ --- as opposed to an $inc$ function that would not.

@@

  \inferrule
  {eq : P = P' \\ red : P' \reduce{i} Q}
  {\constr{struct} \; eq \; red : P \reduce{i} Q}
\end{mathpar}
\caption{Operational semantics indexed by the channel over which reduction occurs.}
\label{fig:reduction}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Resource-aware Typing System}
\label{type-system}

\todo{why do we use a single relation for both type judgements and usage judgments}

@@

\subsection{Multiplicities and Capabilities}
\label{multiplicities}

@@

\subsection{Typing with Leftovers}
\label{leftover typing}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subject Reduction}
\label{subject-reduction}

@@

\subsection{Subject Congruence}
\label{subject-congruence}
